\section{Conclusion}
\label{sec:conclusion}

As semi-autonomous agentic systems move from experimental demonstrations into sustained research use, they expose a structural misalignment between capability and accountability. Current systems can orchestrate multi-step workflows and generate substantial research artifacts, yet the connective infrastructure required to persist, trace, and cumulatively extend those artifacts across time remains underdeveloped. The synthesis presented in this paper identified a consistent divide: existing approaches either maintain structured knowledge without autonomous reasoning, or reason autonomously without enforceable provenance. None yet supports cumulative, provenance-enforced research progression across sessions and documents.

We proposed a provenance-first architectural framework that integrates persistent knowledge graphs with multi-agent orchestration, treating span-level evidence alignment and canonicalization auditing as architectural invariants rather than post-hoc annotations. Together with a faithfulness-centered evaluation framework and a reproducibility protocol for agentic research, this approach reframes provenance from a reporting guideline into foundational research infrastructure. By enforcing structured continuity between claims and evidence over time, the architecture enables not merely isolated task automation, but cumulative, machine-mediated knowledge evolution grounded in traceable sources.

This work does not claim to resolve the broader challenges of autonomous scientific discovery. The proposed system remains bounded by its source literature and the capabilities of its underlying models, and it deliberately retains human oversight to balance autonomy with verifiability. However, if semi-autonomous agents are to participate meaningfully in long-horizon research programs, they must operate within infrastructure that supports persistence, auditability, and structured extension of prior work. Establishing provenance-enforced connective infrastructure is therefore not an optional enhancement, but a prerequisite for accountable, cumulative AI-assisted research ecosystems.